---
title: Introduction to ggplot2
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

## Background

While R provides ready-to-go tools for basic data visualization, it can quickly become cumbersome trying to customize these to more specific needs. The R package _ggplot2_ provides an alternative "grammar of graphics" that makes constructing high-quality, nuanced graphics a lot easier. At the same time, _ggplot2_ accomplishes this by adopting its own syntax and structure, so it takes a little getting used to. It's well worth it though, as I hope this module will show.

## Objectives

We really have one objective today: make a pretty plot! But we can break that down a bit:

  1. Install and load ggplot2
  2. Discuss basics of ggplot syntax
  3. Build a plot
  4. Save the graphic to a file
  
## Install and Load ggplot2

To install an R package, use the ```install.packages()``` function.

```{r echo = TRUE, eval = FALSE}
install.packages("ggplot2")
```

Installing the package gets it onto our computer, but not loaded into our current R session. To do that, we need ```library()```.

```{r echo= TRUE}
library(ggplot2)
```

## Basics of ggplot Syntax

### What's in a ggplot graphic?

The flexibility and elegance of ggplot comes in part from its ["layered grammar of graphics"](http://vita.had.co.nz/papers/layered-grammar.pdf). In essence, ggplot allows us to build complex graphics from multiple "layers" of information. The resulting graphics objects have the following components:

  1. data: the data you want to visualize
  2. mapping: how data are assigned to different visual "aesthetics" -- x-axis, y-axis, shape of points, etc...
  3. geom: a geometric object used to represent the data -- points, bars, lines, etc...
  4. scaling: how values of variables are translated to their visual spacing
  5. faceting: representing subsets of data using subplots

### How do I build one?

The structure of a ggplot graphic also gives us a nice workflow for building one of our own:

  1. specify the data using the ```ggplot()``` function.
  2. add aesthetic mapping using the ```aes()``` function.
    - you can set aesthetics within the ```ggplot()``` function to have them apply to all subsequent layers, or you can set them within each individual ```geom_``` function you use to create the objects representing the data.
  3. Define the geometric objects you'd like to represent your data using the various ```geom_``` functions. These are added _after_ we define the data and any global aesthetic mapping in the ```gplot()``` function. We simply add them to the end of our ggplot workflow with a ```+```. Some example geoms are:
  
  - ```geom_points()```: represents the data with points (requires at least x, y coordinates)
  - ```geom_boxplot()```: represents the data with boxplots (requires at least one variable x)
  - ```geom_violin()```: represents the data with violin plots (requires at least one variable x)
  - ```geom_histogram()```: represents the frequency of different values of variables (requires at least one variable x)
  - ```geom_line()```: represents the data with a line -- helpful for time series. Careful-- not the same as fitting a line to the data.
  - ```geom_smooth()```: represents the data with a smoothed curve and optional confidence intervals. There are options to change what statistical method is used to generate the curve, including linear models, loess regressions, and generalized additive models.

Each type of geom can represent different aesthetics. Points, for example, have a position, color, shape, and size that we can use to represent data. Boxes in a boxplot, on the other hand, have heights, widths, and colors. It's often helpful to check the help file for whatever geom you'd like to use to see what aesthetics are available to map variables to.
    
  4. Add any additional plot elements (e.g., other data subsets, other response variables, trend lines) by appending more ```geom_``` or ```stat_``` functions to the end with ```+```. There are a lot of different options available, and you can even write your own functions to use.
  5. Define scaling of variables if needed (e.g., adjust color range used to represent a variable)
  6. Make adjustments to "fine tune" your graphic -- axes, legends, etc.
  
### Building a Plot

We'll demonstrate this workflow by building a scatterplot, step-by-step. Scatterplots show the relationship between two continuous variables.

#### Load some data

First, we'll need to load some data. As someone who really enjoys playing around with real data, I thought we could work with some published data on Antarctic penguins. These data were collected by Kristen Gorman with the Palmer Station Long Term Ecological Research Program and later developed into an R package for educational uses by Allison Horst, and Alison Hill, and Kristen Gorman. The published manuscript focused on differences in sexual size dimorphism among 3 species of penguin and their relation to sex differences in foraging ecology. The full data set has really neat info on reproductive success and blood measurements of stable isotopes, but for the sake of brevity we will focus on the morphological measures used to quantify sexual size dimorphism.

Let's install the _palmerpenguins_ package.

```{r eval = FALSE, echo = TRUE}
install.packages("palmerpenguins") # install the package
```

And now let's load it.

```{r eval = TRUE, echo = TRUE}
library(palmerpenguins) # load it in current R session
```

The data we'll be using are now available in an object called ```penguins```. The full data are in ```penguins_raw```, which I encourage you to explore on your own later. 

Let's take a look at our data set using the ```print()``` function.

```{r echo = TRUE}
print(penguins)
```

We see that we have variables (columns) for species, island, 4 different morphological variables, sex, and the year of observation. R also gives us some information about this object -- it says that it is a ```tibble```, tells us the size of the tibble (344 rows by 8 columns), and under each variable we get a nice summary of the type of data stored there (factors ("fctr"), machine precision numbers ("dbl"), integers ("int), etc.). The ```tibble``` is a particular type of ```data.frame``` that is used as part of the tidyverse approach to data analysis and visualization. _ggplot2_ is part of the tidyverse, but we can also use it as a standalone package without getting into all the other features of the tidyverse. This is what we'll do here. If you're interested in learning more about tidyverse, check out [this module](TidyverseTest.md).

Let's start by turning the ```tibble``` into a regular ```data.frame```. This isn't strictly necessary because _ggplot2_ would largely treat the data the same way, but we'll do this anyway to formally sever our ties to the rest of the tidyverse. 

```{r echo = TRUE}
penguins_df = as.data.frame(penguins) # Create a new object called penguins_df. This object is the penguins tibble, but as a data frame.
```

Let's check that this did what we intended before we move on. One of the nastier elements of base R's ```data.frame``` is that if we use our ```print()``` function from above to look at it, it will flood the console with data. To get a nice little preview like the tibble gave us, use ```head()```.

```{r echo = TRUE}
head(penguins_df) # use the head() function to look at the first several rows of the penguins_df data frame. Careful that you are now using penguins_df!
```

Great, that looks like the data made it through unscathed.

#### Deciding what to plot

Now we need to decide what we want our plot to look like. Based on an exploration of these data I did in the [module on tidyverse](TidyverseTest.md), I would be interested to know how a penguin's mass scales with its body size, and whether this differs between males and females. I could represent this with a scatterplot, where body size (e.g., flipper length) is on the x-axis and body mass is on the y-axis, and where the points are colored or shaped differently depending on whether the penguin is male or female. It would also be great to add a trend line to plots for males and females to see if the relationship seems different for the sexes, but we'll get to that later.

Try making a sketch of this plot on paper or a whiteboard _before_ you start trying to build it in R. I never cease to surprise myself with how often I _think_ I know what I want to plot, only to find out later that I cannot, in fact, convey my message with that type of plot. While exploring is a necessary part of the process, having a clear mental image of your goal at the outset is invaluable for guiding that journey.

#### Define the data and aesthetic mapping

Now that we know what data we want to plot, and which variables we'd like where, we can tell R about this using the ```ggplot()``` function.

But, if you're like me, you've already forgotten what those variables were named in the data set. Use ```names()``` to refresh your memory first.

```{r echo = TRUE}
names(penguins_df)
```

Okay, now I know I want ```flipper_length_mm``` on the x-axis and ```body_mass_g``` on the y-axis. Eventually, we'll want to color/shape those points by ```"sex"```, but we'll come back to that in a moment.

Let's inform ```ggplot()``` of our intention:

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g)) # data = penguins_df, aes(x = ..., y = ...) defines which variables are mapped to which aesthetics
```

Okay, so we got our x and y axes, and they're appopriately named, but there are no data represented. What gives? Well, we haven't yet added a ```geom_``` function to tell ggplot how to represent those data with a geometric object. 

#### Step 2: Define the geometric object(s)

We'd like these geometrics objects to be points for our scatterplot, so let's use ```geom_point()```. Remember that we add this _after_ the ```ggplot()``` function, using a ```+```. *Make sure that the ```+``` comes _before_ you start a new line with the return key. This tell R that you're not done with ggplot yet, so it will expect more ggplot functions on the next line.

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g)) + # notice the plus sign here *before* the return to start a new line
  geom_point()
```

Hey, now those are some data! Note that we didn't have to specify the aesthetics again inside the ```geom_point()``` function because we set them up for all the layers in the plot using the ```ggplot()``` function. We could have also specified the aesthetics inside ```geom_point()```, for example if we wanted to plot different variables on the same plot. 

Now that we have some points, it's clear that we could show the data by sex if we could just map the points' shape or color to the "sex" variable in our data set. Let's return to our aesthetic mapping to take advantage of this opportunity by using the "color = " argument to the aesthetics.

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, color = sex)) + # add ", color = sex" to the aesthetic mapping in the ggplot() function
  geom_point()
```

Well isn't that nice. Okay, so we have our data, our aesthetic mapping (now with sex as a color!), and the geometric objects (points) representing the data. I don't think I want to add any other geometric objects for now (lines, bars, more points, etc.), so we're ready to make adjustments to the scaling if we'd like.

#### Define scaling of variables if needed

The colors above are okay, but you might want some control over this. The way variables are mapped to aesthetics like size, shape, and color can be controlled through additional ```scale_``` functions like ```scale_color_manual()``` and ```scale_shape_manual()```. These get pretty deep pretty quickly -- there are, for example, several different ```scale_color_``` functions that have different behaviors, and it takes a while to get a good grasp on them.

Let's try using a different color scale. First we will build our own scale using colors we choose based on some identifier (e.g., a name, hex code, rgb value, or number). Then, we'll briefly check out how to use existing color palettes to pick a set of colors at once.

We can change the color of the points in our scatterplot above using ```scale_color_manual()```, provided we tell the function what values of color we'd like to have. We can use (and mix!) syntax for the colors themselves. See the available colors using ```colors()```. Pick a color by name (in quotes), number (e.g., colors()[1]), RGB value, or hex code (also in quotes).

```{r echo = TRUE}
colors()
```

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, color = sex)) +
  geom_point() + 
  scale_color_manual(values = c("tomato2", rgb(0.2, 0.3, 0.6))) # add scale_ function -- I used a name and a set of rgb() values to specify 2 colors
```

That's handy, but you can spend a long time sampling colors and trying to assemble a color palette. Luckily R has lots of built in "color palettes" that we can call on to generate a set of colors to use. These can be divided into palettes roughly analogous to RGB color space (called hue-saturation-value, or HSV colors) and those based on human color perception (hue-chroma-luminance, or HCL colors). I don't have a strong opinion on these at present, but I will focus on HCL here.

What color palettes are available in R if we want to use the HCL color space? We can check using the function ```hcl.pals()```.

```{r echo = TRUE}
hcl.pals()
```

Ooooh, lots of options! Let's pick one and try to see if we can get a set of colors from it. I'm going to try "Harmonic", because that sounds like a cool name. Use the ```hcl.colors()``` function to specify the number of colors you need (n = ...) and the palette you'd like them from (palette = ...). Try one (or a few) of your own!

```{r echo = TRUE}
hcl.colors(n = 2, palette = "Harmonic")
```

So this gives us hex codes for our colors. Nice! We can enter these manually into our ```scale_``` function, or we can nest the ```hcl.colors()``` function inside the ```scale_``` function to skip a step.

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, color = sex)) + # add ", color = sex" to the aesthetic mapping in the ggplot() function
  geom_point() + 
  scale_color_manual(values = hcl.colors(n = 2, palette = "Harmonic"))

# the code above is equivalent to running the code below. The below uses the hex codes.
# ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, color = sex)) + # add ", color = sex" to the aesthetic mapping in the ggplot() function
#   geom_point() + scale_color_manual(values = c("#C7A76C", "#7DB0DD"))
```

Nice! So this might not be my favorite color palette, but now we know how to create our own or use one of the many existing ones.

Notice that ggplot won't apply a color to instances where the variable mapped to the color is NA (see the greyed out points?).

Notice also that ggplot is warning you that there are missing values that are dropped during plotting (i.e., their x or y coordinates are unknown). If we want to formally address this, we can add the ```na.rm = TRUE``` option to our ```ggplot()``` function (affects all layers) or to the corresponding ```geom_```.

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, color = sex)) + # add ", color = sex" to the aesthetic mapping in the ggplot() function
  geom_point(na.rm = TRUE) + 
  scale_color_manual(values = hcl.colors(n = 2, palette = "Harmonic"))
```

Let's try mapping sex to a point shape, too, since you will sometimes need to do this (e.g., for printing in greyscale)

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, shape = sex)) + # put ", shape = sex" in the aesthetic mapping in the ggplot() function
  geom_point(na.rm = TRUE)
```

Finally, we can combine shapes and colors to show even more information (though there may be clearer ways to do this-- see faceting later!). Let's keep shapes as representative of sex, but let colors vary according to species. Note that we'll now need 3 colors, one for each of the 3 species.

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, shape = sex, color = species)) + # put ", shape = sex" in the aesthetic mapping in the ggplot() function
  geom_point(na.rm = TRUE) +
  scale_color_manual(values = hcl.colors(n = 3, palette = "Harmonic"))
```

You'll notice that when we mapped sex to the shape aesthetic for our points, the NAs for sex don't get assigned a symbol type. This means they don't show up in the plot, but NA still shows up in the legend for some reason. That's pretty annoying. In regular practice, this wouldn't really be an issue because we typically would have already filtered our data to remove observations where sex was unknown. However, our present issue gives us an opportunity to explore how to adjust other aspects of our plots like the axes and legends.

#### Make adjustments to "fine tune" your graphic -- axes, legends, etc.

##### Legends

Let's take care of that legend first-- it's driving me crazy! We can adjust the legend using the ```breaks =``` argument inside the corresponding ```scale_``` function for our aesthetic. For example, here we were using sex as a shape. We are getting three discrete "breaks" in sex -- female, male, and NA, but we only want female and male. So we need to use something like ```scale_shape_discrete(breaks = c("female", "male"))```.

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, shape = sex, color = species)) +
  geom_point(na.rm = TRUE) + 
  scale_color_manual(values = hcl.colors(n = 3, palette = "Harmonic")) +
  scale_shape_discrete(breaks = c("female", "male"), )
```

Much better. We changed the legend as a workaround to a problem you shouldn't encounter too often, but you'll often need to modify them as part of your plotting anyway.

Don't like those symbol types? Use ```scale_shape_manual()``` to set them using the ```values = ``` argument like we did with ```scale_color_manual()``` above. This is often necessary when you have a lot of groups, because ```scale_shape_discrete()``` only generates up to 6 and then starts recycling them! Try typing ?pch to see some of the shape types available in R.

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, shape = sex, color = species)) +
  geom_point(na.rm = TRUE) + 
  scale_color_manual(values = hcl.colors(n = 3, palette = "Harmonic")) +
  scale_shape_manual(breaks = c("female", "male"), values = c(10, 4))
```

##### Axis Labels

This one is easy. To specify the x-axis label, just add ```+ xlab("YOUR LABEL HERE")``` to your plot code. I bet you can guess how to change the y-axis label now, too!

```{r echo = TRUE}
ggplot(penguins_df, aes(x = flipper_length_mm, y = body_mass_g, shape = sex, color = species)) +
  geom_point(na.rm = TRUE) + 
  scale_color_manual(values = hcl.colors(n = 3, palette = "Harmonic")) +
  scale_shape_manual(breaks = c("female", "male"), values = c(10, 4)) +
  xlab("Flipper Length (mm)") +
  ylab("Body Mass (g)")
```

Now this is starting to look like a publication-quality graphic! How exciting.

