---
title: Working with Data In Tidyverse
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
  
---

## Background ##

A lot of your time working in R will likely be spent organizing, cleaning, and subsetting data to prepare it for plotting and statistical analysis. The [tidyverse](https://www.tidyverse.org/) is a set of R packages that make this easier by establishing a shared set of standards for how to represent and work with data sets. It includes commonly used packages for managing data like [dplyr](https://dplyr.tidyverse.org/) and [tidyr](https://tidyr.tidyverse.org/), and the [ggplot2](https://ggplot2.tidyverse.org/) graphics package. This module focuses on data management with dplyr and tidyr. For more an introduction to ggplot2, check out [this module](GettingStartedggplot2.md).

## Objectives

The goal of this module is to introduce R users to the philosophy of tidyverse and illustrate how its features can make data management and analysis easier. More specifically, this module guides users through

1. Installing tidyverse
2. Subsetting data
3. Chaining operations using pipes (%>%)
4. Calculating new variables from existing ones
5. Summarizing data using split-apply-combine approach
6. Moving between long- and wide-format data

## Installing Tidyverse

First thing's first -- let's install tidyverse!

```{r eval = FALSE, echo = TRUE}
install.packages("tidyverse") # install the package
```

Note that installing a package just gets it onto our computer-- not into R! To access a package in R, we have to load the package using the `library()` function.

``` {r, eval = TRUE, echo = TRUE}
library(tidyverse) # load it in the current R session.
```

You can see that loading tidyverse actually loads a set of other packages -- _ggplot2_, _purrr_, _tibble_, etc. These are the workhorses of the tidyverse. You can also see a list of "conflicts". These are cases where a package we just loaded (e.g., _dplyr_) has its own version of a base R function (e.g., filter, lag), so it's good to be mindful of this if we use those functions.

## Subsetting data using _dplyr_

The package _dplyr_ within the _tidyverse_ family has a lot of helpful functions for getting your data ready to visualize and analyze. I'll focus on the most commonly used ones in this module, but there's a great cheatsheet available [here](https://github.com/rstudio/cheatsheets/blob/master/tidyr.pdf).

### Commonly used functions

  - select(): select columns
  - filter(): filter rows matching some criteria
  - mutate(): create new columns by applying functions to existing columns
  - group_by(): split data into groups based on one or more variables
  - summarize(): calculate summary statistics for a variable
  - arrange(): sort rows by some criteria
  - count(): count discrete values
  - 
left_join(), right_join(), inner_join(), full_join(): merge data tables in various ways.

*Note*: one of the unique (and handy!) features of tidyverse is that all of these functions return a data frame (what _tidyverse_ calls a "tibble"). In base R, data frames are ```data.frame```s, and subsetting them can return objects that are _not_ still a data.frame. 

### A little help from some penguins

Let's try our hand at some "data wrangling" using some published data on Antarctic penguins. These data were collected by Kristen Gorman with the Palmer Station Long Term Ecological Research Program and later developed into an R package for educational uses by Allison Horst, and Alison Hill, and Kristen Gorman. The published manuscript focused on differences in sexual size dimorphism among 3 species of penguin and their relation to sex differences in foraging ecology. The full data set has really neat info on stable isotope blood measurements and reproductive success, but for the sake of brevity we will focus on the morphological measures used to quantify sexual size dimorphism.

First let's install the _palmerpenguins_ package.

```{r eval = FALSE, echo = TRUE}
install.packages("palmerpenguins") # install the package
```

And now let's load it.

```{r eval = TRUE, echo = TRUE}
library(palmerpenguins) # load it in current R session
```



The data we'll be using are now available in a ```data.frame``` called ```penguins```. The full data are in ```penguins_raw```, which I  encourage you to check out later!

Let's take a look at our data set. The most straightforward way to do this is with the ```print()``` function in base R.

```{r echo = TRUE}
print(penguins)
```
Interesting! It tells us that our dataset is saved in R as a ```tibble```, not a ```data.frame```. *Tibbles are one of the main features of the tidyverse approach to data wrangling.* A ```tibble``` differs from the ```data.frame``` of base R in a few important ways. For one, a ```tibble``` won't drown your console in data when you use the View() function like we did above. But more importantly, a ```tibble``` makes fewer assumptions about what you mean when you build it or mess with it. For example, operations on a ```tibble``` always return a ```tibble``` unless you demand otherwise. In contrast, operations on a ```data.frame``` can return things that are not themselves another ```data.frame```. This is one of the main reasons the tidyverse is so tidy-- your data structure is preserved unless you explicitly ask otherwise. The ```penguins``` dataset comes as a ```tibble``` already because the developers are into the tidyverse phiolosophy.

> - You can change a ```data.frame``` to a ```tibble``` using the  ```as.tibble()``` function.
> - You can change a ```tibble``` to a ```data.frame``` using the ```as.data.frame()``` function.
> - You can create a ```tibble``` from scratch using the ```tibble()``` function [the syntax is largely similar to data.frame(), but with a few handy differences].

Getting back to the data, we see that we have variables for species, island, 4 different morphological variables, sex, and the year of observation. Let's explore some of the tools that _dplyr_ gives us for working with these data.

### select() a column (or columns) of interest

If we're interested in sexual dimorphism of penguins, we'll certainly need to visualize and analyze some of those morphological variables. But how do we perform operations on a single variable within this data set that has many variables? We can use the ```select()``` function to subset the data.

Let's look at body mass first, since this seems to be the most direct measure of overall size.

```{r echo = TRUE}
select(penguins, body_mass_g) ## select(tibble, column)
```
Note that this is *still a tibble*, even though it's just one string of numbers. If we had instead isolated a single variable using the subsetting in base R, we would have gotten a ```vector```, not another ```tibble```. If obtaining the variable as a ```vector``` is the goal, we can accomplish that in _dplyr_ using the ```pull()``` function. The tidyness of the tidyverse comes from the fact that we have to be explicit about this-- in the tidyverse, R will not make assumptions about reformatting our data when we subset it.

Compare below
```{r echo = TRUE}
penguins$"body_mass_g" ## subsetting by variable using base R's $ notation
```

Okay, but back to sexual dimorphism... what we really want to do is to look at body mass _by sex_. Can we get a ```tibble``` with both of those variables?

```{r echo = TRUE}
select(penguins, body_mass_g, sex) ## select(tibble, column)
```

Sure! We can add as many of the variables in the original ```tibble``` to the ```select()``` function's arguments to retain them in our selection.

### filter() rows based on some criteria

